Approximation as Differentiation

### 2025-07-10 main/models.tex
[ tex ] more discussion of semantics GPS in Fam(LatGal)

\todo{Fix a notation for the forwards and backwards maps}

Spelt out in full, it has as objects $(X, \partial X)$, all pairs of a
set $X$ and and for every $x \in X$, a bounded lattice
$\partial X(x)$, and as morphisms
$(X, \partial X) \to (Y, \partial Y)$, all pairs $(f, \partial f)$ of
a function $f: X \to Y$ and for every $x \in X$, a Galois connection
$\partial f(x): \partial X(x) \to \partial Y(f(x))$. The reindexing
along $f$ in the composition
$(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial
  g}{f} \comp \partial f)$ selects the appropriate lattice of
approximations and Galois connection at each point, using the baseline
(unapproximated) function $f$.

@@

\begin{remark}
  \label{rem:further-structure}
  These examples highlight a potential criticism of $\Fam(\LatGal)$ as
  a category for modelling \GPS. For $\mathrm{shortCircuitOr}$, we had
  $(\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top)
  = (\top, \bot)$, indicating that the second argument was not needed
  for computing the output. However, there is no way, in
  $\Fam(\LatGal)$, of turning this into a rigorous statement that the
  $\Set$-component of this morphism does not actually depend on its
  second argument. We conjecture that this can be rectified by
  requiring some kind of additional structure on each object
  $(X, \partial X)$ consisting of a map
  $\Pi_{x : X}.\partial X(x) \to \mathcal{P}(X)$, where $\mathcal{P}$
  is the powerset, which identifies for each $x$ which elements are
  indistinguishable from $x$ at this level of approximation. One would
  also presumably have to require additional conditions for this to
  act like an equivalence relation, and to respect the lattice
  structure\footnote{This additional structure is reminiscent of the
    additional structure on \emph{directed containers} defined by
    \citet{ahman-uustalu}. They require a map
    $\Pi_{x : X}.\partial X(x) \to X$ picking out a specific $X$
    ``jumped to'' by some change $\delta x : \partial X(x)$ at $x$. In
    our proposed setup, we follow the approximation methodology of
    \GPS by having a \emph{set} of things that could be taken instead
    of the original $x$.}.
\end{remark}

\subsubsection{An Approximation Object and the Tagged Data Monad}

The object $\mathbb{A} = (1, \lambda \_.\, \{\top,\bot\})$ represents
a basic object for measuring approximations in terms of presence and
absence. (If all we care about is presence or absence in our
approximations, we could see this object as the analogue of the real
numbers for normal differentiation.)

This object is a commutative monoid in $\Fam(\LatGal)$, and so we can
construct the writer monad $T(X) = \mathbb{A} \times X$ in
$\Fam(\LatGal)$.

On objects $X$ with trivial approximations, we have $\Lift(X) \cong T(X)$,
but for more structured objects we get usage information at every level.

This is a bit like using the writer monad and Moggi's CBN translation
as a basic way of doing provenance tracking, but here we can track
backwards and forwards.

@@

\bob{Describe how \exref{intervals-and-maxima-elements} fits into
  $\Fam(\LatGal)$}

>>>>

Spelt out in full, $\Fam(\LatGal)$ has as objects $(X, \partial X)$,
all pairs of a set $X$ and and for every $x \in X$, a bounded lattice
$\partial X(x)$. Morphisms $(X, \partial X) \to (Y, \partial Y)$, are
triples $(f, \partial f_f, \partial f_r)$ of functions $f : X \to Y$
and families of monotone maps
$\partial f_f : \Pi_{x : X}.\partial X(x) \multimap \partial Y(f\,x)$
(``forward derivative'') and
$\partial f_r : \Pi_{x : X}. \partial Y(f\,x) \multimap \partial X(x)$
(``reverse derivative''), such that for all $x$,
$\partial f_r(x) \dashv \partial f_f(x)$.

@@

\begin{remark}
  \label{rem:further-structure}
  These examples highlight a potential criticism of $\Fam(\LatGal)$ as
  a category for modelling \GPS. For $\mathrm{shortCircuitOr}$, we had
  $(\partial \mathrm{shortCircuitOr})_r(\mathsf{tt},\mathsf{tt})(\top)
  = (\top, \bot)$, indicating that the second argument was not needed
  for computing the output. However, there is no way, in
  $\Fam(\LatGal)$, of turning this into a rigorous statement that the
  $\Set$-component of this morphism does not actually depend on its
  second argument. We conjecture that this can be rectified by
  requiring some kind of additional structure on each object
  $(X, \partial X)$ consisting of a map
  $\Pi_{x : X}.\partial X(x) \to \mathcal{P}(X)$, where $\mathcal{P}$
  is the powerset, which identifies for each $x$ which elements are
  indistinguishable from $x$ at this level of approximation. One would
  also presumably have to require additional conditions for this to
  respect the lattice structure and be preserved by
  morphisms\footnote{This additional structure is reminiscent of the
    additional structure on \emph{directed containers} defined by
    \citet{ahman-uustalu}. They require a map
    $\Pi_{x : X}.\partial X(x) \to X$ picking out a specific $X$
    ``jumped to'' by some change $\delta x : \partial X(x)$ at $x$. In
    our proposed setup, we follow the approximation theme of \GPS by
    having a \emph{set} of things that could be taken instead of the
    original $x$. Objects of $\Fam(\LatGal)$ are ``directed'' in this
    sense because the approximation map can pick out the element of
    the original poset that was approximating $x$.}.
\end{remark}

The $\Lift$ monad provides a controllable way of adding
presence/absence approximation points to composite data, and its monad
structure makes explicit in the program structure exactly how such
approximations are propagated through computations. The fact that
there are different choices of this kind of approximation tracking
provides freedom to the language implementor to decide what
information is worth tracking. The Galois slicing implementations
discussed in \citet{perera12a} and \citet{ricciotti17}, for example,
bake-in an approximation point at every composite type constructor. We
will see in \secref{cbn-translation} that this choice can be
systematised in our setting by considering a monadic CBN translation
to uniformly add $\Lift$ approximation points to composite data types.

\subsubsection{An Approximation Object and the Tagging Monad}

The $\Lift$ monad provides a way of tagging first-order data with
presence and absence information. The object $\Lift(1)$, the lifting
of the terminal object in $\Fam(\LatGal)$, yields an object that
consists purely of presence/absence information:
$\mathbb{A} = (1, \lambda\_.\,\{\top,\bot\})$. This object is the
carrier of a commutative monoid in $\Fam(\LatGal)$, where the forward
maps take the meet (both the inputs are required for the output to be
present) and the backwards maps duplicate.% \bob{Is it also a comonoid?}

Since $\mathbb{A}$ is a monoid, we can define the writer monad
$T(X) = \mathbb{A} \times X$ in $\Fam(\LatGal)$ which ``tags'' $X$
values with approximation information. This is similar to the $\Lift$
monad in that it adds approximation information to an object. On
discrete objects it agrees with the lifting:
$T(\Disc(X)) \cong L(\Disc(X))$. However, on composite data, the two
monads give different approximation lattices. Let $A$ and $B$ be sets.
Then $T(T(\Disc(A)) \times T(T(\Disc(B))))$ has approximation lattices
at $(a,b)$ that are always isomorphic to $\{\top,\bot\}^3$. The
corresponding $\Lift(\Lift(\Disc(A)) \times \Lift(\Disc(B)))$ object's
approximation lattice at $(a,b)$ is always isomorphic to
$(\{\top,\bot\}^2)_\bot$.
In terms of usage tracking, the object using the $\Lift$ monad is more
appealing. The approximation lattice resulting from the use of the $T$
monad contains apparently nonsensical elements corresponding to
``using'' one or other components of the product \emph{without using
  the product itself}. (This arises from the fact that we can project
the $X$ out of $\mathbb{A} \times X$ without touching the
$\mathbb{A}$.)

This example shows that we have to be careful about how we choose the
interpretation of approximable sets in $\Fam(\LatGal)$, and again
highlights the point we made in \remref{further-structure} that
perhaps $\Fam(\LatGal)$ does not have quite enough structure to
determine ``sensible'' approximation information. On the other hand,
the use of the $T$ monad does have the advantage that the
approximation lattices built from discrete sets, products, and
coproducts, are always Boolean lattices, meaning that we can take
complements of usage information. The ability to take boolean
complements of approximations has been used by \citet{perera22} to
compute \emph{related} outputs, as we discuss in
\secref{related-work:galois-slicing}.

@@

So far, the approximation lattices we have looked at in
$\Fam(\LatGal)$ have only consisted of those constructed from finite
products and lifting, and only track binary usage/non-usage
information. \exref{intervals-and-maxima-elements} shows how we can go
beyond this to get more ``quantitative'' approximation
information. Let the object of reals with interval approximations in
$\Fam(\LatGal)$ be
$(\mathbb{R}, \lambda x.\,\{[l,u] \mid l \leq x \leq u\} \cup
\{\bot\})$ where the lattices of intervals are reverse ordered by
inclusion with $\bot$ at the bottom. Then, following the examples in
\exref{intervals-and-maxima-elements} and \exref{stable-functions}, we
can define addition, negation, and scaling by $r \geq 0$:
\begin{displaymath}
  \begin{array}{lcl}
    \mathrm{add}&=&(\begin{array}[t]{@{}l}
      \lambda (x_1,x_2).\,x_1+x_2, \\
      \lambda (x_1,x_2)\,([l_1,u_1],[l_2,u_2]).\,[(l_1+x_2) \sqcap (l_2+x_1),(u_1+x_2) \sqcup (u_2+x_1)] \\
      \lambda (x_1,x_2)\,[l,u].\,([l-x_2,u-x_2],[l-x_1,u-x_1]))
    \end{array}\\
    \mathrm{neg}&=&(\lambda x.\,-x, \lambda x\,[l,u].\,[-u,-l], \lambda x\,[l,u].\,[-u,-l]) \\
    \mathrm{scale}(r)&=&(\lambda x.\,rx, \lambda x\,[l,u].\,[rl,ru], \lambda x\,[l,u].\,\mathrm{if}\,r=0\,\mathrm{then}\,\bot\,\mathrm{else}\,[\frac{l}{r},\frac{u}{r}])
  \end{array}
\end{displaymath}
Scaling by negative numbers is also possible with swapping of lower
and upper bounds, as is multiplication.
